The effect of habitat choice on evolutionary rescue in subdivided populations

### 2020-07-08 ms/main.tex
fin premiere grosse revision main text

\begin{abstract}
    \noindent \textbf{Abstract} Evolutionary rescue is the process by which a declining population successfully adapts genetically\footnote{\francois{remove genetically}\florence{ok to remove}} to avoid extinction. In a structured environment that deteriorates patch by patch, dispersal can substantially alter the chances of evolutionary rescue of a \linelabel{R2-1}\chg{population whose wild type is} not viable in deteriorated patches. Here, we investigate the effect of different dispersal schemes and intensities on the probability of successful establishment of a mutant population adapted to the deteriorated environment. \linelabel{R1-2}\chg{We assume that local fitness is determined by a single haploid locus\florence{ and that d}\st{. D}ispersal is genotype-dependent\st{ and linked to the adaptive trait, i.e. dispersal does not evolve by itself}\footnote{\florence{[otherwise it sounds like $m$ is changing...] [need to add that it's the bias that is changing, not the probability to emigrate]}}. In this scenario, we }find that the probability of evolutionary rescue can undergo up to three phases when increasing the rate of dispersal\footnote{\florence{since you specify ``at low disp'', ``at intermediate disp'' etc., you can remove ``when increasing the rate of dispersal'' to save words}}: (i) at low dispersal rates, the probability of establishment of a mutant population increases; (ii) at intermediate dispersal rates, the establishment probability decreases; (iii) at large dispersal rates, the population homogenizes, \florence{, which either promotes or suppresses}\footnote{\florence{old version: either promoting or suppressing}} the process of evolutionary rescue, depending on the fitness difference between the mutant and the wild type. Our results show that habitat choice\footnote{\florence{up to now, the reader does not know this is about habitat choice (``schemes'' is not precise enough}}, when compared to \chg{unbiased} dispersal, impedes successful adaptation when the mutant has the same habitat preference as the wild type, but promotes adaptation when the mutant mainly immigrates into patches where it has a growth advantage over the wild type.
\end{abstract}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Evol rescue and empirical evidence %%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Current anthropogenic environmental changes such as deforestation, soil and water contamination or rising temperatures, contribute to the decline of the populations of many species, that might eventually go extinct \citep{bellard_2012, diniz_2019}. Pests and pathogens experience similarly strong selective pressures as a result of increased consumption of antibiotics and use of pesticides \citep{ramsayer_2013, kreiner_2018}. 
The process of genetic\footnote{\florence{rescue by plasticity as well}} adaptation that saves populations from extinction is termed evolutionary rescue. This \linelabel{R1-3}\chg{process is characterized by an initial population decline (that would result in population extinction) followed by recovery due to the establishment of adapted genotypes, } \florence{classically} resulting in a U-shaped demographic trajectory over time~\citep{gomulkiewicz_1995}.
In recent years, empirical examples of evolutionary rescue have accumulated \citep[as reviewed by ][]{alexander_2014,carlson_2014,bell_2017}. Laboratory experiments have provided direct evidence of evolutionary rescue \linelabel{R1-4} \citep[e.g.][]{bell_2009, agashe_2011, lachapelle_2012, lindsey_2013, stelkens_2014}. In the wild, however, demographic and genotypic data are rarely monitored together at the same time, which impedes direct observation of evolutionary rescue. Still, evolutionary rescue has been suggested as a mechanism that has saved a few wild populations from extinction \citep[e.g.][]{vanderwal_2012, digiallonardo_2015, gignoux_2018}.

@@

\florence{[Need a paragraph on the genetic makeup. We are not explaining this at all at the moment and this is lacking] Rescue can occur through plasticity (refs Chevin? \footnote{https://royalsocietypublishing.org/doi/full/10.1098/rspb.2016.1690 , https://royalsocietypublishing.org/doi/abs/10.1098/rstb.2012.0089 and/or https://onlinelibrary.wiley.com/doi/full/10.1111/j.1558-5646.2009.00875.x}) or genetic adaptation (refs). The traits involves can be continuous (ref) or discrete (ref). In this work, we consider genetic adaptation mediated by discrete traits. Specifically, we consider that a single mutation can prevent extinction, and...}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Evol rescue and dispersal %%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
In mathematical models, evolutionary rescue is often studied in a spatially homogeneous situation where the whole population experiences a sudden decrease in habitat quality. In this setting, a large number of theoretical results have been established, for example on the effects of recombination \citep{uecker_2015} and 
horizontal gene transfer \citep{tazzyman_2014}, 
reproduction mechanisms \citep{glemin_2013,uecker_2017}, intra- and interspecific competition \citep{osmond_2013}, predation pressure \linelabel{R1-5}\citep{yamamichi_2015}, 
bottlenecks \citep{martin_2013}, 
different genetic pathways \citep{osmond_2019}, and the context-dependent fitness effects of mutations \citep{anciaux_2018}. \chg{In contrast to these abrupt change scenarios, evolutionary rescue can also be studied in a gradually changing environment \citep[e.g.][]{osmond_2017}}\footnote{\pete{Does anybody know other references?} \hildegard{What about Lynch et al. 1991. Adaptive and  demographic responses of plankton populations to environmental change. Limnology and Oceanography 36:1301--1312., B{\"u}rger and Lynch 1995. Evolution and extinction in a changing environment: a quantitative genetic analysis. Evolution 49:151--163 and Lande and Shannon 1996. The role of  genetic  variation  in adaptation and population persistence in a changing environment. Evolution 50:434--437? But please check; it's a long time that I read those...}}. 
Such gradual changes can in particular occur in fragmented environments.

In fragmented environments, habitat deterioration is not necessarily synchronized across patches: there can be a transient spatially heterogeneous environment consisting of a mosaic of old and of degraded habitat patches, until eventually the whole environment has deteriorated. If individuals that populate different patches are able to move between those, the effect of dispersal on evolutionary rescue needs to be taken into account \citep{uecker_2014, tomasini_2019}. \linelabel{R1-1}\chg{The intensity of dispersal among patches tunes how abruptly environmental change is experienced. With very low dispersal, patches are essentially isolated from each other, and each patch undergoes an abrupt change independently of the other patches. With higher dispersal, asynchronous deterioration among patches is experienced as a more gradual change overall.}
Experiments that study the effect of dispersal on evolutionary rescue are rare, but, for instance, \citet{bell_2011} found that \linelabel{R1-6} dispersal can increase the likelihood of successful genetic adaptation.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Adaptation and dispersal %%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The transient mosaic of degraded and non-degraded patches that results from asynchronous degradation in a fragmented habitat is similar to the setting of models of source-sink dynamics. These models represent a spatially heterogeneous environment, constant in time, in which wild-type populations in unfavorable habitats can only be maintained thanks to recurrent immigration from favorable habitats. 
Experimental and theoretical studies have found that \linelabel{R2-3}\chg{increasing} dispersal can have a positive or a negative effect on genetic adaptation in a heterogeneous environment (see e.g., for studies on discrete traits, \florence{[refocus the citations with papers on discrete traits; otherwise there are too many possible citations -- it's fine to do so given that we now explicitly say the paper is focused on discrete traits]}\citet{holt_1997,gomulkiewicz_1999} for positive; \citet{storfer_1998,garcia_1997,kirkpatrick_1997,fedorka_2012} for negative; and \linelabel{R1-7}\chg{\citet{kawecki_2000, gallet_2018}} for both effects).
%Reason for different results: (i) demography or not (Kawecki); (ii) selection strength (Gallet), (iii) selection strength (Ronce)

%%% DISPERSAL, biased or not
In theoretical studies of local adaptation and evolutionary rescue, dispersal is typically assumed to be \chg{unbiased}, i.e. dispersing individuals are distributed uniformly among patches. Only few investigations in the context of local adaptation in source-sink systems have taken into account non-\chg{uniform} dispersal patterns \citep[e.g.][]{kawecki_1995,holt_1996,kawecki_2002,amarasekare_2004}. This analytical focus on \chg{unbiased} dispersal is in stark contrast to dispersal schemes observed in nature \citep[][]{edelaar_2008,clobert_2009,edelaar_2012}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Empirical examples for non-random dispersal %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
One of the best documented modes of non-\chg{uniform} dispersal is density-dependent dispersal. Density dependence can be positive or negative: either individuals prefer to settle or stay in large groups (positive density-dependence), or they choose to remain in or move to less populated regions (negative density-dependence). Density-dependent dispersal, of either form, is ubiquitously found in nature and has been reported in many species across the tree of life, including insects \citep{endriss_2019}, spiders \citep{meester2010}, amphibians \citep{gautier_2006}, birds \citep{wilson_2017}, fishes \citep{turgeon_2012}, and mammals~\citep{stoen_2006}.

Another well-established dispersal scheme is a type of habitat choice, whereby individuals tend to immigrate into habitats they are best adapted to. This mechanism has for example been reported in lizards \citep{bestion_2015}, birds \citep{dreiss_2011,benkman_2017}, fishes \citep{bolnick_2009}, worms \citep{mathieu_2010}, and ciliates \citep{jacob_2017,jacob_2018}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% What do we do? %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We model an environment that consists of various patches with one of two possible habitats: the `old' habitat, in which both \linelabel{R1-9-1}\chg{types, wild type and mutant, have a positive growth rate}, and the `new' habitat, where in the absence of immigration the wild-type population will eventually go extinct. We study four biologically motivated dispersal schemes, which correspond to the four combinations of biases towards old vs.\ new patches for wild type and mutants, and we compare these dispersal schemes to \chg{unbiased} dispersal. Our analysis is carried out step-wise. We first consider a temporally constant but spatially heterogeneous environment with two (`old' and `new') patch types. In this setting, we first study the probability of establishment of a single mutant, assuming there are no further mutations between types. We then relax the assumption of no further mutations, and compute a probability of adaptation, i.e. of establishment of the mutant lineage\footnote{\florence{not sure about this defn, please check}}. Finally, we let habitat degradation proceed, assuming that patches, one after \chg{another}, deteriorate over time until all locations contain the new habitat. Using the previous results, we approximate a probability of evolutionary rescue, i.e.\ that a mutation appears, establishes, thereby allowing the population to persist in spite of environmental degradation. We find that dispersal biases affect the probabilities of establishment and of evolutionary rescue, because it directly affects the local growth rates.

@@

\subsection*{Main assumptions and life-cycle}
%%%%%%%%%%%%%%%%%%%%%%%%%%
% General description %%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%
We consider a spatially structured environment consisting of $M$ patches all connected to each other. The habitat of a patch is either in the \textit{old} or in the \textit{new} state, corresponding to habitat quality before and after environmental deterioration, respectively. One after \chg{another} every $\tau$ generations, the habitat of a patch deteriorates, from old to new state, the transition being irreversible. Initially ($t<0$), all patches are of the old-habitat type. At time $t=0$, the first patch deteriorates. After $(M-1)\tau$ generations, all patches are of the new-habitat type. We denote the time-dependent frequency of old-habitat patches by $f_{\text{old}}$. It equals $1$ before the first environmental change takes place ($t<0$), and decreases by $1/M$ after each environmental deterioration event, until it eventually hits~$0$, when all patches have undergone the environmental change. This setting corresponds to the one analyzed by \citet{uecker_2014}, and more recently by \citet{tomasini_2019} in the special case of just two patches\footnote{\florence{what about carrying capacities in these papers?}}. The maximum numbers of individuals that can live in a patch of a given habitat type, i.e.\ the carrying capacities, are denoted $K_{\text{old}}$ and $K_{\text{new}}$ for old- and new-habitat patches respectively; $K_{\text{old}}$ and $K_{\text{new}}$ may differ.

The population living in this environment consists of asexually reproducing, haploid individuals; generations are discrete and non-overlapping. There are two possible types of individuals, wild types and mutants. The individuals go through the following life-cycle: \begin{enumerate}[label = (\roman*)]
	\item Dispersal: individuals may move between patches. Further details about this step are given below.
	\item Reproduction: individuals reproduce within patches. \linelabel{R2-7}The number of offspring produced by an individual of type $i$ in habitat $k$ (before density regulation, if any), i.e.\ its fecundity, is drawn in a Poisson distribution of expectation $\omega_i^k$. Having fewer than $1$ offspring in expectation means that the local subpopulation will get extinct in the absence of immigration, because the deaths of the parents at each generation are not compensated by enough births on average. On the contrary, the local population is viable if the expected fecundity is greater than $1$. We assume that wild-type and mutant populations are viable in old-habitat patches, and that the mutant's expected fecundity there is lower than the wild type's: $1 \leq \omega^{\text{old}}_m < \omega^{\text{old}}_w$ \footnote{\florence{In the table, it is $0 \leq $, yet the populations are supposed to be viable}}. In new-habitat patches, a wild-type population will eventually go extinct, while a mutant one would persist, hence the term ``rescue mutant'': $\omega^{\text{new}}_w < 1 < \omega^{\text{new}}_m$. 
	All parents die at the end of this step.
	\item \linelabel{R2-9} Mutation: wild-type offspring mutate to the rescue mutant type with probability $\theta$ (back mutations from the mutant to the wild type are neglected).
	\item \linelabel{R2-8}Regulation: if the number of offspring produced locally exceeds the local carrying capacity $K_k$ (where $k$ refers to the habitat type, old or new), the population size is down-regulated to the $K_k$: individuals are randomly removed until the local population size is equal to $K_k$. One mutant offspring has the same chance of being removed as one wild-type offspring: \linelabel{AE-3} we assume that wild-type and mutant individuals are competitively equivalent. If the number of offspring is below the carrying capacity, the regulation step is ignored. We call ``successful offspring'' offspring that survive the density regulation step, and become adults at the next generation. 
	At the end of this step, all offspring become adults, and a new cycle then begins.
\end{enumerate}

@@

The probability that an individual of type $i$ migrates from a patch of any habitat type into a patch of the new-habitat type is\florence{TODO change into simple bias (remove m)}
%
\begin{equation}\label{eq:dispersal}
\begin{aligned}
m_i^{\text{new}} &= m\, \frac{1-f_{\text{old}}}{1-f_{\text{old}}+\chg{e^{\pi_i}} f_{\text{old}}} = 1-m_i^{\text{old}},\\
\end{aligned}
\end{equation}
\florence{there seems to be a problem with the second equation, should be $m -$ no? }
where, as defined above, $f_{\text{old}}$ is the frequency of old-habitat patches and $\pi_i$ the dispersal bias into old-habitat patches. The use of an exponential $e^{\pi_i}$ \linelabel{R1-19} ensures that the fraction in eq.~\eqref{eq:dispersal} is positive and between zero and one. For the ease of notation we will write $\widehat{\pi}_i=e^{\pi_i}$.\footnote{\florence{TODO remove hat notation}}

@@

& + \frac{m}{\sqrt{C}}\left( a_{\text{new}}(1-f_{\text{old}}) + a_{\text{old}} \widehat{\pi}_m f_{\text{old}} - (a_{\text{new}}-a_{\text{old}}) e^{\pi_m} f_{\text{old}} \right)\, ,

@@

\begin{figure}[t!]
	\centering
	\begin{subfigure}{.5\textwidth}
  		\centering
  		\includegraphics[width=\linewidth]{fig2a.pdf}
  		\caption{$\varphi_{\text{old}}$ with $\omega^\text{old}_m = 1.35$ (large fecundity difference)}
  		\label{fig:vary_m_est_old-large}
	\end{subfigure}%
	\begin{subfigure}{.5\textwidth}
 		 \centering
 		 \includegraphics[width=\linewidth]{fig2b.pdf}
  		\caption{$\varphi_{\text{new}}$ with $\omega^\text{old}_m = 1.35$}
  		\label{fig:vary_m_est_new-large}
	\end{subfigure}
	\begin{subfigure}{.5\textwidth}
  		\centering
  		\includegraphics[width=\linewidth]{fig2c.pdf}
  		\caption{$\varphi_{\text{old}}$ with $\omega^\text{old}_m = 1.45$ (small fecundity difference)}
  		\label{fig:vary_m_est_old-small}
	\end{subfigure}%
	\begin{subfigure}{.5\textwidth}
 		 \centering
 		 \includegraphics[width=\linewidth]{fig2d.pdf}
  		\caption{$\varphi_{\text{new}}$ with $\omega^\text{old}_m = 1.45$}
  		\label{fig:vary_m_est_new-small}		
	\end{subfigure}
	\caption{\textbf{Establishment probability when varying the emigration rate.} \small We plot the theoretical results for $\varphi_{\text{old}}$ in (a) and (c), and for $\varphi_{\text{new}}$ in (b) and (d) for $\omega^\text{old}_m=1.35$ in (a), (b) and $\omega^\text{old}_m=1.45$ in (c), (d). Comparison with the results from stochastic simulations (symbols) show very good agreement with our approximation found in eq.~\eqref{eq:estab_approx} (dashed lines). The solid lines are the numerical solution of eq.~\eqref{eq:ext_prob}. The lines in (a) show a clear separation of the three regions of the establishment probability as discussed in the main text: {\color{change}(i) an initial increase due to higher probabilities of dispersal from old- to new-habitat patches; (ii) a local maximum due to increasing number of mutants emigrating from new-habitat patches; (iii) an increase due to relaxed competition in old-habitat patches, a result of wild-type emigration.}}
	\label{fig:vary_m_est}
\end{figure}

>>>>

\begin{abstract}
	\florence{will do this in the very end}
    \noindent \textbf{Abstract} Evolutionary rescue is the process by which a declining population successfully adapts genetically\footnote{\francois{remove genetically}\florence{ok to remove}} to avoid extinction. In a structured environment that deteriorates patch by patch, dispersal can substantially alter the chances of evolutionary rescue of a \linelabel{R2-1}\chg{population whose wild type is} not viable in deteriorated patches. Here, we investigate the effect of different dispersal schemes and intensities on the probability of successful establishment of a mutant population adapted to the deteriorated environment. \linelabel{R1-2}\chg{We assume that local fitness is determined by a single haploid locus\florence{ and that d}\st{. D}ispersal is genotype-dependent\st{ and linked to the adaptive trait, i.e. dispersal does not evolve by itself}\footnote{\florence{[otherwise it sounds like $m$ is changing...] [need to add that it's the bias that is changing, not the probability to emigrate]}}. In this scenario, we }find that the probability of evolutionary rescue can undergo up to three phases when increasing the rate of dispersal\footnote{\florence{since you specify ``at low disp'', ``at intermediate disp'' etc., you can remove ``when increasing the rate of dispersal'' to save words}}: (i) at low dispersal rates, the probability of establishment of a mutant population increases; (ii) at intermediate dispersal rates, the establishment probability decreases; (iii) at large dispersal rates, the population homogenizes, \florence{, which either promotes or suppresses}\footnote{\florence{old version: either promoting or suppressing}} the process of evolutionary rescue, depending on the fitness difference between the mutant and the wild type. Our results show that habitat choice\footnote{\florence{up to now, the reader does not know this is about habitat choice (``schemes'' is not precise enough}}, when compared to \chg{unbiased} dispersal, impedes successful adaptation when the mutant has the same habitat preference as the wild type, but promotes adaptation when the mutant mainly immigrates into patches where it has a growth advantage over the wild type.
\end{abstract}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Evol rescue and empirical evidence %%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Current anthropogenic environmental changes such as deforestation, soil and water contamination or rising temperatures, contribute to the decline of the populations of many species, that might eventually go extinct \citep{diniz_2019}. Pests and pathogens experience similarly strong selective pressures as a result of increased consumption of antibiotics and use of pesticides \citep{ramsayer_2013, kreiner_2018}. 
The process of adaptation that saves populations from extinction is termed evolutionary rescue. This \linelabel{R1-3}process is characterized by an initial population decline (that would result in population extinction) followed by recovery due to the establishment of adapted genotypes, classically resulting in a U-shaped demographic trajectory over time~\citep{gomulkiewicz_1995}.
In recent years, empirical examples of evolutionary rescue have accumulated \citep[as reviewed by ][]{alexander_2014,carlson_2014,bell_2017}. Laboratory experiments have provided direct evidence of evolutionary rescue \linelabel{R1-4} \citep[e.g.][]{bell_2009, agashe_2011, lachapelle_2012, lindsey_2013, stelkens_2014}. In the wild, however, demographic and genotypic data are rarely monitored together at the same time, which impedes direct observation of evolutionary rescue. Still, evolutionary rescue has been suggested as a mechanism that has saved a few wild populations from extinction \citep[e.g.][]{vanderwal_2012, digiallonardo_2015, gignoux_2018}.

@@

\florence{[Need a paragraph on the genetic makeup. We are not explaining this at all at the moment and this is lacking] Rescue can occur through plasticity (refs Chevin? \footnote{https://royalsocietypublishing.org/doi/full/10.1098/rspb.2016.1690 , https://royalsocietypublishing.org/doi/abs/10.1098/rstb.2012.0089 and/or https://onlinelibrary.wiley.com/doi/full/10.1111/j.1558-5646.2009.00875.x}) or genetic adaptation (refs). The traits involves can be continuous (ref) or discrete (ref). In this work, we consider genetic adaptation mediated by a discrete trait. }

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Evol rescue and dispersal %%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
In mathematical models, evolutionary rescue is often studied in a spatially homogeneous situation where the whole population experiences a sudden decrease in habitat quality. In this setting, a large number of theoretical results have been established, for example on the effects of recombination \citep{uecker_2015} and 
horizontal gene transfer \citep{tazzyman_2014}, 
reproduction mechanisms \citep{glemin_2013,uecker_2017}, intra- and interspecific competition \citep{osmond_2013}, predation pressure \linelabel{R1-5}\citep{yamamichi_2015}, 
bottlenecks \citep{martin_2013}, 
different genetic pathways \citep{osmond_2019}, and the context-dependent fitness effects of mutations \citep{anciaux_2018}. In contrast to these abrupt change scenarios, evolutionary rescue can also be studied in a gradually changing environment \citep[e.g.][]{osmond_2017}\footnote{\pete{Does anybody know other references?} \hildegard{What about Lynch et al. 1991. Adaptive and  demographic responses of plankton populations to environmental change. Limnology and Oceanography 36:1301--1312., B{\"u}rger and Lynch 1995. Evolution and extinction in a changing environment: a quantitative genetic analysis. Evolution 49:151--163 and Lande and Shannon 1996. The role of  genetic  variation  in adaptation and population persistence in a changing environment. Evolution 50:434--437? But please check; it's a long time that I read those...}}. 
Such gradual changes can in particular occur in fragmented environments.

In fragmented environments, habitat deterioration is not necessarily synchronized across patches: there can be a transient spatially heterogeneous environment consisting of a mosaic of old and of degraded habitat patches, until eventually the whole environment has deteriorated. If individuals that populate different patches are able to move between those, the effect of dispersal on evolutionary rescue needs to be taken into account \citep{uecker_2014, tomasini_2019}. \linelabel{R1-1}The intensity of dispersal among patches tunes how abruptly environmental change is experienced. With very low dispersal, patches are essentially isolated from each other, and each patch undergoes an abrupt change independently of the other patches. With higher dispersal, asynchronous deterioration among patches is experienced as a more gradual change overall.
Experiments that study the effect of dispersal on evolutionary rescue are rare, but, for instance, \citet{bell_2011} found that \linelabel{R1-6} dispersal can increase the likelihood of successful genetic adaptation.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Adaptation and dispersal %%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The transient mosaic of degraded and non-degraded patches that results from asynchronous degradation in a fragmented habitat is similar to the setting of models of source-sink dynamics. These models represent a spatially heterogeneous environment, constant in time, in which wild-type populations in unfavorable habitats can only be maintained thanks to recurrent immigration from favorable habitats. 
Experimental and theoretical studies have found that \linelabel{R2-3}increasing dispersal can have a positive or a negative effect on genetic adaptation in a heterogeneous environment (see e.g., for studies on discrete traits, \florence{[refocus the citations with papers on discrete traits; otherwise there are too many possible citations -- it's fine to do so given that we now explicitly say the paper is focused on discrete traits]}\citet{holt_1997,gomulkiewicz_1999} for positive; \citet{storfer_1998,garcia_1997,kirkpatrick_1997,fedorka_2012} for negative; and \linelabel{R1-7}\chg{\citet{kawecki_2000, gallet_2018}} for both effects).
%Reason for different results: (i) demography or not (Kawecki); (ii) selection strength (Gallet), (iii) selection strength (Ronce)

%%% DISPERSAL, biased or not
In theoretical studies of local adaptation and evolutionary rescue, dispersal is typically assumed to be unbiased, i.e. dispersing individuals are distributed uniformly among patches. Only few investigations in the context of local adaptation in source-sink systems have taken into account non-uniform dispersal patterns \citep[e.g.][]{kawecki_1995,holt_1996,kawecki_2002,amarasekare_2004}. This analytical focus on unbiased dispersal is in stark contrast to dispersal schemes observed in nature \citep[][]{edelaar_2008,clobert_2009,edelaar_2012}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Empirical examples for non-random dispersal %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
One of the best documented modes of non-uniform dispersal is density-dependent dispersal. Density dependence can be positive or negative: either individuals prefer to settle or stay in large groups (positive density-dependence), or they choose to remain in or move to less populated regions (negative density-dependence). Density-dependent dispersal, of either form, is ubiquitously found in nature and has been reported in many species across the tree of life, including insects \citep{endriss_2019}, spiders \citep{meester2010}, amphibians \citep{gautier_2006}, birds \citep{wilson_2017}, fishes \citep{turgeon_2012}, and mammals~\citep{stoen_2006}.

Another well-established dispersal scheme is a type of habitat choice whereby individuals tend to immigrate into habitats they are best adapted to. This mechanism has for example been reported in lizards \citep{bestion_2015}, birds \citep{dreiss_2011,benkman_2017}, fishes \citep{bolnick_2009}, worms \citep{mathieu_2010}, and ciliates \citep{jacob_2017,jacob_2018}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% What do we do? %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We model an environment that consists of various patches with one of two possible habitats: the `old' habitat, in which both \linelabel{R1-9-1}types, wild type and mutant, have a positive growth rate, and the `new' habitat, where in the absence of immigration the wild-type population will eventually go extinct. We study four biologically motivated dispersal schemes, which correspond to the four combinations of biases towards old vs.\ new patches for wild type and mutants, and we compare these dispersal schemes to unbiased dispersal. Our analysis is carried out step-wise. We first consider a temporally constant but spatially heterogeneous environment with two (`old' and `new') patch types. In this setting, we first study the probability of establishment of a single mutant, assuming there are no further mutations between types. We then relax the assumption of no further mutations, and compute a probability of adaptation, i.e. of establishment of the mutant lineage\footnote{\florence{not sure about this defn, please check}}. Finally, we let habitat degradation proceed, assuming that patches, one after another, deteriorate over time until all locations contain the new habitat. Using the previous results, we approximate a probability of evolutionary rescue, i.e.\ that a mutation appears, establishes, thereby allowing the population to persist in spite of environmental degradation. We find that dispersal biases affect the probabilities of establishment and of evolutionary rescue.

@@

\subsection*{Main assumptions and life-cycle}
%%%%%%%%%%%%%%%%%%%%%%%%%%
% General description %%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%
We consider a spatially structured environment consisting of $M$ patches all connected to each other. The habitat of a patch is either in the \textit{old} or in the \textit{new} state, corresponding to habitat quality before and after environmental deterioration, respectively. One after another every $\tau$ generations, the habitat of a patch deteriorates, from old to new state, the transition being irreversible. Initially ($t<0$), all patches are of the old-habitat type. At time $t=0$, the first patch deteriorates. After $(M-1)\tau$ generations, all patches are of the new-habitat type. We denote the time-dependent frequency of old-habitat patches by $f_{\text{old}}$. It equals $1$ before the first environmental change takes place ($t<0$), and decreases by $1/M$ after each environmental deterioration event, until it eventually hits~$0$, when all patches have undergone the environmental change. This setting corresponds to the one analyzed by \citet{uecker_2014}, and more recently by \citet{tomasini_2019} in the special case of just two patches\footnote{\florence{what about carrying capacities in these papers?}}. The maximum numbers of individuals that can live in a patch of a given habitat type, i.e.\ the carrying capacities, are denoted $K_{\text{old}}$ and $K_{\text{new}}$ for old- and new-habitat patches respectively; $K_{\text{old}}$ and $K_{\text{new}}$ may differ.

The population living in this environment consists of asexually reproducing, haploid individuals; generations are discrete and non-overlapping. There are two possible types of individuals, wild types and mutants. The individuals go through the following life-cycle: \begin{enumerate}[label = (\roman*)]
	\item Dispersal: individuals may move between patches. Further details about this step are given below.
	\item Reproduction: individuals reproduce within patches. \linelabel{R2-7}The number of offspring produced by an individual of type $i$ in habitat $k$ (before density regulation, if any), i.e.\ its fecundity, is drawn in a Poisson distribution of expectation $\omega_i^k$. Having fewer than $1$ offspring in expectation means that the local subpopulation will get extinct in the absence of immigration, because the deaths of the parents at each generation are not compensated by enough births on average. On the contrary, the local population is viable if the expected fecundity is greater than $1$. We assume that wild-type and mutant populations are viable in old-habitat patches\footnote{\florence{this won't be the case in fig \ref{fig:rescue_lethal}}}, and that the mutant's expected fecundity there is lower than the wild type's: $1 \leq \omega^{\text{old}}_m < \omega^{\text{old}}_w$ \footnote{\florence{In the table, it is $0 \leq $, yet the populations are supposed to be viable}}. In new-habitat patches, a wild-type population will eventually go extinct, while a mutant one would persist, hence the term ``rescue mutant'': $\omega^{\text{new}}_w < 1 < \omega^{\text{new}}_m$. 
	All parents die at the end of this step.
	\item \linelabel{R2-9} Mutation: wild-type offspring mutate to the rescue mutant type with probability $\theta$ (back mutations from the mutant to the wild type are neglected).
	\item \linelabel{R2-8}Regulation: if the number of offspring produced locally exceeds the local carrying capacity $K_k$ (where $k$ refers to the habitat type, old or new), the population size is down-regulated to the $K_k$: individuals are randomly removed until the local population size is equal to $K_k$. One mutant offspring has the same chance of being removed as one wild-type offspring: \linelabel{AE-3} we assume that wild-type and mutant individuals are competitively equivalent. If the number of offspring is below the carrying capacity, the regulation step is ignored. We call ``successful offspring'' offspring that survive the density regulation step, and become adults at the next generation. 
	At the end of this step, all offspring become adults, and a new cycle then begins.
\end{enumerate}

@@

The probability that a dispersing individual of type $i$ settles in a patch of the new-habitat type is
%
\begin{equation}\label{eq:dispersal}
\begin{aligned}
\mu_i^{\text{new}} &= \frac{1-f_{\text{old}}}{1-f_{\text{old}}+ e^{\pi_i} f_{\text{old}}} = 1 - \mu_i^{\text{old}},\\
\end{aligned}
\end{equation}
where, as defined above, $f_{\text{old}}$ is the frequency of old-habitat patches and $\pi_i$ the dispersal bias into old-habitat patches. The use of an exponential $e^{\pi_i}$ \linelabel{R1-19} ensures that the fraction in eq.~\eqref{eq:dispersal} is positive and between zero and one.

@@

& + \frac{m}{\sqrt{C}}\left( a_{\text{new}}(1-f_{\text{old}}) + a_{\text{old}} e^{\pi_m} f_{\text{old}} - (a_{\text{new}}-a_{\text{old}}) e^{\pi_m} f_{\text{old}} \right)\, ,

@@

\begin{figure}[t!]
	\centering
	\begin{subfigure}{.5\textwidth}
  		\centering
  		\includegraphics[width=\linewidth]{fig2a.pdf}
  		\caption{$\varphi_{\text{old}}$ with $\omega^\text{old}_m = 1.35$}
  		\label{fig:vary_m_est_old-35}
	\end{subfigure}%
	\begin{subfigure}{.5\textwidth}
 		 \centering
 		 \includegraphics[width=\linewidth]{fig2b.pdf}
  		\caption{$\varphi_{\text{new}}$ with $\omega^\text{old}_m = 1.35$}
  		\label{fig:vary_m_est_new-35}
	\end{subfigure}
	\begin{subfigure}{.5\textwidth}
  		\centering
  		\includegraphics[width=\linewidth]{fig2c.pdf}
  		\caption{$\varphi_{\text{old}}$ with $\omega^\text{old}_m = 1.45$}
  		\label{fig:vary_m_est_old-45}
	\end{subfigure}%
	\begin{subfigure}{.5\textwidth}
 		 \centering
 		 \includegraphics[width=\linewidth]{fig2d.pdf}
  		\caption{$\varphi_{\text{new}}$ with $\omega^\text{old}_m = 1.45$}
  		\label{fig:vary_m_est_new-45}		
	\end{subfigure}
	\caption{\textbf{Establishment probability as a function of the emigration rate.} \small Panels (a) and (c) show the establishment probabilities when the first mutant arises in an old-habitat patch ($\varphi_{\text{old}}$), and panels (b) and (d) the establishment probabilities when the mutant arises in a new-habitat patch ($\varphi_{\text{new}}$), for mutant fecundity in old-habitat patches $\omega^\text{old}_m=1.35$ in (a), (b) and $\omega^\text{old}_m=1.45$ in (c), (d). Markers: simulations; full lines: numerical solution of eq.~\eqref{eq:ext_prob}; dashed lines, approximate solution shown eq.~\eqref{eq:estab_approx}. The colors and marker shapes correspond to the different dispersal schemes, with the same parameters as in Fig.\ref{fig:disp_schemes}.}
	\label{fig:vary_m_est}
\end{figure}
